A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-06-28 calculus.tex
Evaluation
Progress on example
Clean up the equations a bit

We use the convention of splitting the formulas into linear (in)equalities
($\SomeInequalities$) and other formulas ($\SomeClause$), and assume that
predicates $\Image$ and $\Connected$ only occur positively. Since our rules
operate by adding and matching linear (in)equalities in a proof goal, we use the
shorthand of listing the matched inequalities as antecedents e.g., in
\Propagate{}).

@@

\subsection{An Example}
Here we will return to $A'$ from \cref{sec:intuition} and performs the steps of
\cref{sec:intuition:algebra,sec:intuition:prop,sec:intuition:split} with the
formal calculus we just established, but exclude \cref{const:s2-in-b} and
therefore the entire automaton $B$ since we introduce support for products of
automata in \cref{sec:multiple}. Note that this example then becomes
satisfiable, as $s_1 = \Char{dd}, i = 1, s_2 = \Char{adda}, l_A = 4$ is a
satisfying assignment to
\cref{const:more-inside-than-before,const:s1-in-c-dd,const:s1-substring}.

Starting with~$A'$, $\Map$ extracts the increments of a transition (see
\cref{fig:a_1}), $h(\FromLabelTo{q_1}{c / i}{q_2}) = \Vec{v}$ for the increment
vector $\Vec{v}$, concretely $h(\FromLabelTo{I_A}{\Sigma / \left[ l_{A+}, i_+
\right]}{I_A}) = \left[ l_{A+}, i_+ \right]$. We use the same compact notation
here as in \cref{sec:intuition} to represent what is essentially a sparse vector
of $1$ and $0$ coefficients, i.e. = $\left[ l_{A+}, i_+ \right] = \left[1, 1,
0\right]$. Initially, we let $\Filter$ map to fresh variables to obtain:
\begin{equation}
    \begin{aligned}
      \Filter(\FromLabelTo{D}{d/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{F_A}) & = \TransitionVar_1    & 
                \Filter(\FromLabelTo{F_A}{\Sigma/\begin{bmatrix} l_{A+} \end{bmatrix}}{F_A} ) & = \TransitionVar_2 &       \FlowEq(A', \Filter) & = \begin{aligned}
                    1 + \cancel{\TransitionVar_5} &= \cancel{\TransitionVar_5} + \TransitionVar_3 + \TransitionVar_1\\
                    \TransitionVar_4 &= \TransitionVar_1 \\
                    \TransitionVar_3 + \TransitionVar_1 + \cancel{\TransitionVar_2} &\geq \cancel{\TransitionVar_2}  \\
                  \end{aligned} \\
      \Filter(\FromLabelTo{I_A}{c/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{F_A} ) & =  \TransitionVar_3 & 
                \Filter(\FromLabelTo{I_A}{d/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{D}) & = \TransitionVar_4 &\\
      \Filter(\FromLabelTo{I_A}{\Sigma/\begin{bmatrix} l_{A+} i_+\end{bmatrix}}{I_A}) & = \TransitionVar_5 & 
    \end{aligned}
\end{equation}\label{eq:single-filter}

@@

\begin{figure}[ht]
  \centering
\begin{prooftree}
  \hypo[]{{\text{Regular solving finishes the proof}}}
  \infer1[]{
    \begin{aligned}
    i = \TransitionVar_5 \land 
    n = 2 \land \\
    2 \geq i \land l_A - i - 2 > 0 
    \land \TransitionVar_2 = l_A - i - 2 
    \land \TransitionVar_3 = 0 \\
    \end{aligned}}
  \infer1[\Subsume: \cref{sec:intuition:subsume}]{
    \begin{aligned}
    \Connected(A', \Filter) \land 
    \TransitionVar_1 = 1 \land 
    i = \TransitionVar_5 \land 
    n = 2 \land 
    2 \geq i \land l_A - i - 2 > 0  \\
    \land \TransitionVar_2 = l_A - i - 2 
    \land \TransitionVar_3 = 0 \\
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: $\TransitionVar_1 = 1$ \cref{sec:a_1}]{
    \begin{aligned}
    \Connected(A', \Filter) \land \\
    i = \TransitionVar_5 \land 
    n = 2\TransitionVar_1 \land n \geq i \land l_A - i - n > 0 
    \land \TransitionVar_2 = l_A - i - n 
    \land \TransitionVar_3 = 1 - \TransitionVar_1 
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: $=$-elimination]{
    \begin{aligned}
    \Connected(A', \Filter)  
    \land i = \TransitionVar_5 
    \land n = \TransitionVar_4 + \TransitionVar_1 
    \land n \geq i 
    \land l_A - i - n > 0 \\
    \land \TransitionVar_4 = \TransitionVar_1 
    \land \TransitionVar_2 = l_A - i - n 
    \land \TransitionVar_3 = 1 - \TransitionVar_1
    \land 1 = \TransitionVar_3 + \TransitionVar_1 
    \land \TransitionVar_4 = \TransitionVar_1 
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: expand $\Sigma, \FlowEq$]{
    \begin{aligned}
    i = \TransitionVar_5 
    \land l_A = \sum_{k = 1}^{k = 5} \TransitionVar_k 
    \land n \geq i \land l_A - i - n > 0
    \land n = \TransitionVar_4 + \TransitionVar_1\\
    \land \Connected(A', \Filter) 
    \land \FlowEq(A', \Filter)
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: expand sum]{
    \Connected(A', \Filter) \land
    \FlowEq(A', \Filter) \land
    \left[l_A, i, n\right] =
    \sum\limits_{\Transition \in \Transitions_{A'} } \Map(\Transition) \cdot \Filter(\Transition)
  }
  \infer1[\Expand{}]{\Image{}_{A', \Map}(\Filter, \left[l_A, i, n\right]) \land n \geq i \land l_A - i - n > 0}
\end{prooftree}
\caption{A derivation tree for an application of
\cref{const:s1-in-c-dd,const:s1-substring,const:something-before-and-after,const:more-inside-than-before}
from \cref{ex:string-constraints}, corresponding to handling the Parikh
automaton $A'$ of \cref{fig:aut_a}. Note the liberal use of removal of equal
variables (e.g. if $x = y$ appears as a statement, we substitute $y$ for $x$
everywhere and drop $\ldots \land x=y \land \ldots$) to keep the size of the
equations managable.}\label{fig:derivation:single}
\end{figure}

>>>>

We use the convention of splitting the formulas into linear (in)equalities
($\SomeInequalities$) and other formulas ($\SomeClause$), and assume that
predicates $\Image$ and $\Connected$ only occur positively. Since our rules
operate by adding and matching linear (in)equalities in a proof goal, we use the
shorthand of listing the matched inequalities as antecedents e.g., in
\Propagate{}.

@@

\subsection{An Example}\label{sec:single:example}
Here we will return to $A'$ from \cref{sec:intuition} and performs the steps of
\cref{sec:intuition:algebra,sec:intuition:prop,sec:intuition:split} with the
formal calculus we just established, but exclude \cref{const:s2-in-b} and
therefore the entire automaton $B$ since we introduce support for products of
automata in \cref{sec:multiple}. Note that this example then becomes
satisfiable, as $s_1 = \Char{dd}, i = 1, s_2 = \Char{adda}, l_A = 4$ is a
satisfying assignment to
\cref{const:more-inside-than-before,const:s1-in-c-dd,const:s1-substring}.

Starting with~$A'$, $\Map$ extracts the increments of a transition (see
\cref{fig:a_1}), $h(\FromLabelTo{q_1}{c / i}{q_2}) = \Vec{v}$ for the increment
vector $\Vec{v}$, concretely $h(\FromLabelTo{I_A}{\Sigma / \left[ l_{A+}, i_+
\right]}{I_A}) = \left[ l_{A+}, i_+ \right]$. We use the same compact notation
here as in \cref{sec:intuition} to represent what is essentially a sparse vector
of $1$ and $0$ coefficients, i.e. = $\left[ l_{A+}, i_+ \right] = \left[1, 1,
0\right]$. Initially, we let $\Filter$ map to fresh variables to obtain:
\begin{equation}\label{eq:single-filter}
    \begin{aligned}
      & \Filter(\FromLabelTo{D}{d/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{F_A}) = \TransitionVar_1
        & \Filter(\FromLabelTo{F_A}{\Sigma/\begin{bmatrix} l_{A+} \end{bmatrix}}{F_A} ) = \TransitionVar_2   
        & \FlowEq(A', \Filter)  = \begin{aligned}
                       1 + \cancel{\TransitionVar_5} &= \cancel{\TransitionVar_5} + \TransitionVar_3 + \TransitionVar_1\\
                    \TransitionVar_4 &= \TransitionVar_1 \\
                    \TransitionVar_3 + \TransitionVar_1 + \cancel{\TransitionVar_2} &\geq \cancel{\TransitionVar_2}  \\
                    \end{aligned} \\
      &\Filter(\FromLabelTo{I_A}{c/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{F_A} ) =  \TransitionVar_3 
          & \Filter(\FromLabelTo{I_A}{d/\begin{bmatrix}l_{A+}, n_+\end{bmatrix}}{D}) = \TransitionVar_4
          & \Filter(\FromLabelTo{I_A}{\Sigma/\begin{bmatrix} l_{A+} i_+\end{bmatrix}}{I_A}) = \TransitionVar_5  
    \end{aligned}
\end{equation}

@@

\begin{figure}[ht]
  \centering
\begin{prooftree}
  \hypo[]{{\text{Regular solving finishes the proof}}}
  \infer1[\EquationReasoning{}: $i = 1$]{
  i = 1
  \land n = 2
  \land \TransitionVar_2 = l_A - 3
  \land \TransitionVar_3 = 0
  \land \TransitionVar_5 = 1
  \land \TransitionVar_1 = \TransitionVar_4 = 1
  }
  \infer1[]{
    \begin{aligned}
    i = \TransitionVar_5 \land 
    n = 2 \land \\
    0 < i < 2
    \land l_A - i - 2 > 0 
    \land \TransitionVar_2 = l_A - i - 2 
    \land \TransitionVar_3 = 0 \\
    \end{aligned}}
  \infer1[\Subsume: \cref{sec:intuition:subsume}]{
    \begin{aligned}
    \Connected(A', \Filter) \land 
    \TransitionVar_1 = 1 \land 
    i = \TransitionVar_5 \land 
    n = 2 \land \\
    0 < i < 2 
    \land l_A - i - 2 > 0
    \land \TransitionVar_2 = l_A - i - 2 
    \land \TransitionVar_3 = 0
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: $\TransitionVar_1 = 1$ \cref{sec:a_1}]{
    \begin{aligned}
    \Connected(A', \Filter) \land
    i = \TransitionVar_5 \land 
    n = 2\TransitionVar_1 
    \land n > i > 0 \\
    \land l_A - i - n > 0 
    \land \TransitionVar_2 = l_A - i - n 
    \land \TransitionVar_3 = 1 - \TransitionVar_1 
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: $=$-elimination]{
    \begin{aligned}
    \Connected(A', \Filter)  
    \land i = \TransitionVar_5 
    \land n = \TransitionVar_4 + \TransitionVar_1 
    \land n > i > 0
    \land l_A - i - n > 0
    \land \TransitionVar_4 = \TransitionVar_1 \\
    \land \TransitionVar_2 = l_A - i - n 
    \land \TransitionVar_3 = 1 - \TransitionVar_1
    \land 1 = \TransitionVar_3 + \TransitionVar_1 
    \land i > 0
    \land \TransitionVar_4 = \TransitionVar_1 
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: expand $\Sigma, \FlowEq$]{
    \begin{aligned}
    i = \TransitionVar_5 
    \land l_A = \sum_{k = 1}^{k = 5} \TransitionVar_k 
    \land n > i > 0
    \land l_A - i - n > 0
    \land n = \TransitionVar_4 + \TransitionVar_1\\
    \land \Connected(A', \Filter) 
    \land \FlowEq(A', \Filter)
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: expand sum]{
    \Connected(A', \Filter) \land
    \FlowEq(A', \Filter) \land
    \left[l_A, i, n\right] =
    \sum\limits_{\Transition \in \Transitions_{A'} } \Map(\Transition) \cdot \Filter(\Transition)
  }
  \infer1[\Expand{}]{
    \Image{}_{A', \Map}(\Filter, \left[l_A, i, n\right]) 
    \land n \geq i 
    \land l_A - i - n > 0
    \land i > 0
    }
\end{prooftree}
\caption{A derivation tree for an application of
\cref{const:s1-in-c-dd,const:s1-substring,const:something-before-and-after,const:more-inside-than-before}
from \cref{ex:string-constraints}, corresponding to handling the Parikh
automaton $A'$ of \cref{fig:aut_a}. Note the liberal use of removal of equal
variables (e.g. if $x = y$ appears as a statement, we substitute $y$ for $x$
everywhere and drop $\ldots \land x=y \land \ldots$) to keep the size of the
equations managable.}\label{fig:derivation:single}
\end{figure}

### 2023-06-28 evaluation.tex
Clean up the equations a bit

In \cref{fig:solve-division} we show how many of the~\NrBenchmarks{} instances
the respective back-end could solve. A summary of their outcomes by instance
type is also available in \cref{tab:solve-status}. Note that many instances lack
a corum ground truth as they are solved by only one backend. We see that
\Calculus{} generally outperforms \Nuxmv{} on determining unsatisfiability,
while being similar at satisfiable instances. Both \Nuxmv{} and \Calculus{}
outperform baseline on satisfiable instances, and \Calculus{} outperforms
\Nuxmv{}.

Baseline performs worse on satisfiable instances because it executes a heuristic
meant to detect unsatisfiability early at a heavy penalty to satisfiable
instances roughly similar to \cite{approximate-parikh}. The heuristic is enabled
since we suspect that unsatisifiable instances are more common; at least we have
seen more solved unsatisfiable instances.

@@

A cactus plot showing the number of instances solved within a given timeout for
each backend configuration can be seen in \cref{fig:cactus}. Here, we see
clearly that \Calculus{} outperforms \Nuxmv{} in general, but that \Nuxmv{}
might have better long-term scaling properties. Notably, for machines with
better single-core performance \Nuxmv{}'s scaling benefits kicks in faster,
though the graph looks similar in shape.

\iffalse
\subsection{Finding a Presburger Formula}\label{sec:evaluation:finding-image}

For baseline and \Calculus{}, \Catra{} offers the ability to find the equivalent Presburger formula representing a given instance. For baseline, we use the built-in quantifier elimination facilities of the underlying \Princess{} theorem prover, while for \Catra{} we use the specially tailored approach described in \cref{sec:finding-the-image}. For this experiment, we use only the~\NrKnownSat{} instances known to be satisfiable from the previous experiment detailed in \cref{sec:scaling,sec:runtime}.

To make sure baseline puts up as much competition as possible, we disable
checking intermittent satisfiability and configure \Catra{} to run in the
maximally eager mode where the product is first computed before any
satisifiability check is performed. We run the experiments with a timeout
of~\ImageTimeout{}. The results of the experiment is summarised in
\cref{fig:cactus:image} and \cref{tab:image-results}. \Fudge{We see here that
something happens}.

\begin{figure}[ht]
  \caption{The number of instances \Catra{} was able to find the Presburger form of the image for within a given number of seconds per backend.}
  \label{fig:cactus:image}
\end{figure}
\fi

>>>>

In \cref{fig:solve-division} we show how many of the~\NrBenchmarks{} instances
the respective back-end could solve. A summary of their outcomes by instance
type is also available in \cref{tab:solve-status}. Note that many instances lack
a corum ground truth as they are solved by only one backend. We see that
\Calculus{} generally outperforms \Nuxmv{} on determining unsatisfiability, as
does baseline while being similar at satisfiable instances. Both \Calculus{} and
baseline outperforms baseline on satisfiable instances. On satisfiable
instances, \Nuxmv and \Calculus{} has similar performance.

Baseline performs worse on satisfiable instances because it executes a heuristic
meant to detect unsatisfiability early at a heavy penalty to satisfiable
instances roughly similar to \cite{approximate-parikh}. The heuristic is enabled
since the improvement is significant compared to the extra cost and since we
have observed more unsatisfiable instances than satisfiable ones.

@@

A cactus plot showing the number of instances solved within a given timeout can
be seen in \cref{fig:cactus}. We see that \Calculus{} outperforms \Nuxmv{} in
general, but that \Nuxmv{} might have better scaling properties on longer runs.
For machines with better single-core performance than our experiment machine
\Nuxmv{}'s scaling benefits kicks in faster, so it is likely that the amount of
CPU cycles spent on the problem is the underlying factor. This is likely due to
two factors. First, \Nuxmv{} is more mature than \Calculus{}, and it makes sense
that its more general model checking methods would pay off on more difficult
instances where more problem-specific methods offer less payoff. Second, for
longer running calculations clause learning as described in
\cref{sec:clause-learning} can be assumed to matter more. However, the clauses
we learn are still relatively unpolished, and in particular generalising
insights across product computations is nontrivial and therefore partially left
as future work.

### 2023-06-28 multiple.tex
Clean up the equations a bit

We return again to our example in \cref{sec:intuition}.We also reuse the
definition of $\Map$ as the simple projection to extract the counter increments,
so in this case, $\Map(\Transition) = \left[l_{B+}\right]$ for every
transition~$\Transition$ since we are only counting the length. The reader is
strongly advised to review \cref{fig:aut_a_annotated,fig:aut_b_annotated} from
\cref{sec:intuition} while going through this example! In particular, note that
the transition variables of the figures correspond to the ones used here.

The transition variables of \cref{sec:single:example} recur
\cref{eq:single-filter}, but we add the definitions for $B$ shown in
\cref{eq:multi-filter}:
\begin{equation}\label{eq:multi-filter}
    \begin{aligned}
        \Filter(\FromLabelTo{A}{a / l_{B+}}{F_B} ) = \TransitionVar_6\\
        \Filter(\FromLabelTo{A}{c / l_{B+}}{A} ) = \TransitionVar_7\\
        \Filter(\FromLabelTo{B}{b / l_{B+}}{I_B})= \TransitionVar_8 \\
        \Filter(\FromLabelTo{B}{c / l_{B+}}{B})= \TransitionVar_9\\
        \Filter(\FromLabelTo{B}{c / l_{B+}}{F_B})= \TransitionVar_{10}\\
        \Filter(\FromLabelTo{I_B}{a / l_{B+}}{A})= \TransitionVar_{11}\\
        \Filter(\FromLabelTo{I_B}{b / l_{B+}}{B})= \TransitionVar_{12}\\
        \FlowEq(B, \Filter) =  \begin{aligned}
            1 + \TransitionVar_{8}   &= \TransitionVar_{11} + \TransitionVar_{12}\\
            \land \TransitionVar_{11} &= \TransitionVar_6\\
            \land \TransitionVar_{12} &= \TransitionVar_8 + \TransitionVar_{10}
        \end{aligned}
    \end{aligned}
\end{equation}

@@

%\subsubsection{\Calculus{} for products of automata terminates}
\begin{lemma}\label{lma:multi-terminates}
  Suppose $\SomeClause{}$ is a set of formulas in which the product
  version of $\Image$ only occurs positively. There is no
  infinite sequence of proofs~$P_0, P_1, P_2, \ldots$ in which $P_0$
  has $\SomeClause{}$ as root, and each $P_{i+1}$ is derived from
  $P_i$ by applying one of the rules in \cref{tbl:rules:multi}.
\end{lemma}

@@

%\subsubsection{The rules in \cref{tbl:rules:multi} are solution-preserving}

>>>>

We return again to our example in \cref{sec:intuition}. We also reuse the
definition of $\Map$ as the simple projection to extract the counter increments,
so in this case, $\Map(\Transition) = \left[l_{B+}\right]$ for every
transition~$\Transition$ since we are only counting the length. The reader is
strongly advised to review \cref{fig:aut_a_annotated,fig:aut_b_annotated} from
\cref{sec:intuition} while going through this example! In particular, note that
the transition variables of the figures correspond to the ones used here.

The transition variables of \cref{sec:single:example} recur
\cref{eq:single-filter}, but we add the definitions for $B$ shown in
\cref{eq:multi-filter}:
\begin{equation}\label{eq:multi-filter}
    \begin{aligned}
        & \Filter(\FromLabelTo{A}{a / l_{B+}}{F_B} ) = \TransitionVar_6 
        & \Filter(\FromLabelTo{A}{c / l_{B+}}{A} ) = \TransitionVar_7 
        & \Filter(\FromLabelTo{B}{b / l_{B+}}{I_B}) = \TransitionVar_8 
        &\Filter(\FromLabelTo{B}{c / l_{B+}}{B}) = \TransitionVar_9  \\
        & \Filter(\FromLabelTo{B}{c / l_{B+}}{F_B})= \TransitionVar_{10}
        & \Filter(\FromLabelTo{I_B}{a / l_{B+}}{A})= \TransitionVar_{11}
        & \Filter(\FromLabelTo{I_B}{b / l_{B+}}{B})= \TransitionVar_{12}
        \FlowEq(B, \Filter) &=  \begin{aligned}
            1 + \TransitionVar_{8}   &= \TransitionVar_{11} + \TransitionVar_{12}\\
            \land \TransitionVar_{11} &= \TransitionVar_6\\
            \land \TransitionVar_{12} &= \TransitionVar_8 + \TransitionVar_{10}
        \end{aligned}
    \end{aligned}
\end{equation}

@@

\subsubsection{\Calculus{} for products of automata terminates}
\begin{lemma}\label{lma:multi-terminates}
  Suppose $\SomeClause{}$ is a set of formulas in which the product
  version of $\Image$ only occurs positively. There is no
  infinite sequence of proofs~$P_0, P_1, P_2, \ldots$ in which $P_0$
  has $\SomeClause{}$ as root, and each $P_{i+1}$ is derived from
  $P_i$ by applying one of the rules in \cref{tbl:rules:multi}.
\end{lemma}

@@

\subsubsection{The rules in \cref{tbl:rules:multi} are solution-preserving}
